A Mechanised Library for Generic (Co)Programming

### 2024-03-16 paper/paper.tex
minor

\section{Mechanising Extractable Container Functors}
In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
who have a initial algebras. A common approach to construct such functors is to
use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
However, reasoning about container equality
will require us to consider both functional extensionality and heterogeneous
equality. We avoid these axioms by introducing a
custom equivalence relation on types (Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:containers}
Functors are functions \coq{F : Type -> Type} which additionally have a map
\coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
functor represents a container for abstract data that can be manipulated
without changing the outer structure. For example, the type 
\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
over elements of type \coq{A} with the obvious \coq{fmap} function recursive
over a list and applying the function \coq{A -> B} for each element of the
list. Additionally, the type of lists \coq{List A} arises as the least
fixed-point of the functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, and it is not possible to build
the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
positive.  Due to this, we are going to restrict to \emph{polynomial functors}
as represented by containers.  A container is defined by a type of
\emph{shapes} and a \emph{family of position types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would define
\coq{Shape} as \coq{unit + A} indicating there are two constructors in the data
type and one of these contains a piece of data of type \coq{A}.  Moreover, we
would define the positions as \coq{Pos (inl tt) = Empty_set} and 
\coq{Pos (inr a) = unit} indicating that the first constructor does not have
any type variables and the second constructor has one type variable.

At this point an \emph{extension} of this container is a functor defined as
follows:
\begin{coqcode}
Record App (X : Type) := MkApp { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms given by post-composition
with \coq{contents}.
\begin{coqcode}
Definition fmap (f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}
In our running example, the type \coq{App X}, for
some \coq{X} has two inhabitants.  The first is a pair composed
by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
in fact a function of type \coq{Empty_set -> X}. This pair is, therefore
isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
second inhabitant is the pair composed by \coq{inr a : Shape}, for
some \coq{a : A}, and a function \coq{Pos (inr a) -> X}.  This latter type is
equal to \coq{unit -> X} which represents the elements of \coq{X} and therefore
is isomorphic to \coq{X}.  This particular pair of shapes and positions is 
therefore isomorphic to the type \coq{unit + A * X}.

The correspondence between containers and polynomial functors has been
formalised in this work and can be found in the accompanying code (file
Container.v). 
% 
%Generally, any functor that
%is isomorphic to a container extension is called a \emph{container functor}.
%Here we will call \emph{container functor} the functors defined by a specific
%container extension.

\subsection{Extractable Containers}
In the previous section we defined containers using dependent types.  However,
Ocaml's type system is not equipped to handle these. To solve this problem
Coq's code extraction mechanism will insert unsafe casts.

Consider instead representing the positions of a container using a decidable
validity predicate (\coq{valid}) which assigns shapes to positions. We use
boolean functions and coercions from \coq{bool} to \coq{Prop} to represent
decidable predicates, similarly to SSReflect~\cite{GonthierL09}. The family of
positions for a given shape in a container can be recovered by defining:
\begin{coqcode}
Class Cont (Shape : Type) (APos : Type) := { valid : Shape -> APos -> bool }.
Record Pos `(C : Cont Shape APos) (shape : Shape) :=
  ValidPos { val : APos; IsValid : valid shape pos }.
\end{coqcode}
Coq's code extraction will now be able to erase the validity predicate, and
generate OCaml code that is free of unsafe casts. The OCaml code extracted for
\coq{Pos} will now be exactly the code extracted for \coq{APos}.
The decidability of the validity predicate is crucial for our purposes of 
remaining axiom-free. To illustrate this, suppose that we need to show
the equality of two container extensions. We will need to show that, for
the same positions, they will produce the same result. In Coq, the goal
would look as follows: 
\begin{coqcode}
 k : Pos C s -> X
 P1, P2 : valid s p
 ---------------------------
 k (ValidPos p P1) = k (ValidPos p P2)
\end{coqcode}
If \coq{valid} was a regular proposition in \coq{Prop}, it would not possible
to prove the equality of \coq{P1} and \coq{P2}. However, by using a decidable
predicate, if we know that \coq{P1} and \coq{P2} are of type 
\coq{valid s p = true}, then we can prove without any axioms that 
\coq{P1 = P2 = eq_refl}.

@@

\subsubsection{Equality of Container Extensions}
Reasoning about the equality of container extensions is not entirely solved by
using decidable validity predicates to define the families of positions. In
general, we want to equate container extensions that have the same shape, and
that, for equal shape and position, they return the same element. To avoid the
use of the functional extensionality axiom, we capture this relation with the
following inductive proposition in Coq:
\begin{coqcode}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph { app :> A -> B; app_eq : forall x y, x =e y -> app x =e app y }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type \coq{A ~> B} will be
extracted to the OCaml equivalent to \coq{A -> B}. Note the implicit coercion
from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}
Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:algebras}, and the
mechanisation of initial algebras of container extensions.
%The definitions in
%Figure~\ref{fig:poly} are redefined to account for these changes in a
%straightforward way, i.e. we prove that validity predicates are indeed proper
%morphisms.

@@

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}.  Since every morphism \coq{f : A ~> B}
satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
morphism as a proper element of Coq's respectfulness relation. In practice,
this means that we can use the \coq{rewrite} tactic on proofs of type 
\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D) (H : h \o g =e id), 
  h \o (g \o f) =e f.
Proof.  rewrite compA, H.  reflexivity. Qed.
\end{coqcode}

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types.
%These equivalences are
%be used when building proper morphisms that manipulate their respective types.
\begin{coqcode}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{coqcode}
Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\coq{fun x => x} and \coq{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \coq{fun f g x => (f x, g x)},
with the projections being the standard Coq  \coq{fst} and
\coq{snd} functions. Similarly, sum injections are encoded using
Coq's \coq{inl} and \coq{inr} constructors, and
pattern matching on them uses the function 
\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and  flipping the arguments of
a proper morphism. We force most of our definitions to be inlined, to help
Coq's code extraction mechanism to inline as many of these combinators as
possible.

@@

\subsubsection{Example 1}
Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

@@

\subsubsection{Example 2} We then prove that the length function fuses with
the naive quadratic \emph{reverse} function:
\begin{coqcode}
Definition sf2 : Ext (length \o reverse). 
  calculate.  unfold length, reverse. rewrite hylo_fusion_l. 
  2:{ (* Rewrite into the fused version *) }
  simpl; reflexivity.
Defined.
\end{coqcode}
This code extracts to the optimised length function on the input list:
\begin{ocamlcode}
let rec ex2 = function | [] -> (Uint63.of_int (0))
                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
\end{ocamlcode}

>>>>

\section{Mechanising Extractable Container Functors}
\dcas{Subsections are tentative topics to remember what I need to fill in here,
but these are not final}

In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e. those
who have a initial algebras. A common approach to construct such functors is to
use \emph{container}~\cite{AbbottAG05} which is what we introduce in this
Section~\ref{sec:containers}.  However, reasoning about containers equality will
require us to consider both functional extensionality and heterogeneous
equality. To avoid this in Section~\ref{sec:types-morphisms} we introduce a
custom equivalence relation on types.

\subsection{Functors and Containers}
\label{sec:containers}
To abstract away the shape of a particular data structure we use functors.
These are functions \coq{F : Type -> Type} which additionally have a
map \coq{fmap : forall A B, (A -> B) -> F A -> F B} witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursive over a list and applying the
function \coq{A -> B} for each element of the list.
Additionally the type of lists arises as the least fixed-point for the
functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, so we are going to restrict
ourselves to \emph{polynomial functors} as represented by containers.  A
container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and \coq{Pos (inr a) = unit} indicating that the first constructor does not have
any recursive type variables and the second constructor has one type variable.

At this point an \emph{extension} of this container is a functor defined as
follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms given by post-composition
with \coq{contents}.  In our running example, the type \coq{App X}, for
some \coq{X} has two inhabitants.  The first is a pair composed
by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
in fact a function of type we \coq{Empty_set -> X}  by definitions which in turn
is isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
second inhabitant is the pair composed by \coq{inr a : Shape}, for
some \coq{a : A}, and a function \coq{Pos (inr a) -> X}.  This latter type is
equal to \coq{unit -> X} which represents the elements of \coq{X} and therefore
is isomorphic to \coq{X}.  This particular pair of shapes and positions is
isomorphic to the type \coq{unit + A * X}.

The correspondence between containers and polynomial functors has been
formalised in this work and has been moved to the
Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
is isomorphic to a container extension is called a \emph{container functor}.
Here we will call \emph{container functor} the functors defined by a specific
container extension.

\subsection{Extractable Containers}
In the previous section we defined containers using dependent types.  However,
Ocaml's type system is not equipped to handle these and to solve this problem
Coq's code extraction mechanism will insert unsafe casts.

Here consider instead representing the positions of a container using a validity
predicate \coq{Valid : Shape -> Pos -> bool} which assigns shapes to positions.
Moreover, it is crucial for our purposes of remaining axiom-free that this
predicate is \emph{decidable} in order to be able to reason about equality of
container extensions and, as a result, the contents of the container extensions
need to be accessed via a function \coq{contents} whose domain depends on the
shape of the container.  This means that for a given type of shapes and
positions, the equality of two objects of a container extension will need to
deal with heterogeneous equality.
\mpav{This part needs more explanation.}

\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape -> Pos -> bool }.
Coercion is_true (b : bool) : Prop := b = true.
Record PosOf `(C : Cont Shape Pos) (shape : Shape) :=
  ValidPos { val : Pos; IsValid : valid shape pos }.
\end{coqcode}
where \coq{valid} is the \emph{decidable} validity predicate. Note that booleans
can be lifted to propositions similarly to the implementation provided by
SSReflect~\cite{GonthierL09}. This allows us to define the dependent pair of
valid positions as a record which will be extracted to a singleton type since
the validity proof is erased by Coq's code extraction mechanism.

We now define \emph{container extensions} as follows:
\begin{coqcode}
Record App `(C : Cont Shape Pos) (X : Type) :=
  MkApp { shape : Shape; contents : PosOf C shape -> X }.
\end{coqcode}
The functorial action for container extensions on morphisms boils down to
function composition, which trivially preserves identity and composition.
\begin{coqcode}
Definition fmap `(C : Cont Shape Pos) `(f : A -> B) (x : App C A) : App C B
  := {| shape := shape x; contents := fun e => f (contents x e) |}
\end{coqcode}

@@

Finally we can define the type of containers as follows:

\subsubsection{Equality of Container Extensions}
The equality of container extensions needs to be extensional, since the
contents are obtained by applying a function. We want to equate container
extensions that have the same shape, and that, for equal shape and position,
they return the same element. To avoid the use of functional extensionality
axioms, we capture this relation with the following inductive proposition in
Coq:
\begin{coqcode}
Inductive AppR `{F : Cont Sh P} {X} (x y : App F X) : Prop :=
| AppR_ext (Es : shape x = shape y)
           (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 = cont y e2).
\end{coqcode}
Note that  we do not care about the validity proof of the positions, only their
value. This is to simplify (slightly) our proofs. This relation is
trivially \emph{reflexive}, \emph{transitive}, and \emph{symmetric}.

@@

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}. Furthermore, we add a coercion from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way, i.e. we prove that validity predicates are indeed proper
morphisms.  \dcas{Should we show, instead, the actual full definition? It
becomes a bit more complicated ...}

@@

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}.  Since every morphism \coq{f : A ~> B}
satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
morphism as a proper element of Coq's respectfulness relation. In practice,
this means that we can use the \coq{rewrite} tactic on proofs of type 
\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D), 
  h \o g =e id -> h \o (g \o f) =e f.
Proof.
  intros H.  rewrite compA H.  reflexivity.
Qed.
\end{coqcode}

\subsubsection{Polynomial Types}
We define a number of equivalences for polynomial types. These equivalences are
be used when building proper morphisms that manipulate their respective types.
\begin{coqcode}
Instance ext_eq (A : Type) `{eq_B : setoid B} : setoid (A -> B).
Instance pair_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A * B).
Instance sum_eq `{eq_A : setoid A} `{eq_B : setoid B} : setoid (A + B).
Instance prop_eq : setoid Prop.
Instance pred_sub `{eA : setoid A} {P : A -> Prop} : setoid {a : A | P a}.
\end{coqcode}

Most of the definitions that involve functions and polynomial types are
straightworward.  Identity and composition are defined as 
\coq{fun x => x} and \coq{fun f g x => f (g x)} 
respectively, and the proofs that they are proper morphisms is straightforward,
and automatically discharged by Coq. 
Products are built using function \coq{fun f g x => (f x, g x)},
with the projections being the standard Coq  \coq{fst} and
\coq{snd} functions. Similarly, sum injections are encoded using
Coq's \coq{inl} and \coq{inr} constructors, and
pattern matching on them uses the function 
\coq{fun f g x => match x with | inl y => f y | inr y => g y end}.
The proofs that these morphisms are proper are straightforward.  Finally, we
also provide functions for currying/uncurrying, and partially applying a proper
morphism. We avoid using definitions, or unfold them, to help Coq's code
extraction mechanism to inline as many of these small definitions as possible.

@@

Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

@@

We then prove that the length function fuses with
the naive quadratic \emph{reverse} function:
\begin{coqcode}
Definition sf2 : Ext (length \o reverse). 
  calculate.  unfold length, reverse. rewrite hylo_fusion_l. 
  2:{ (* Rewrite into the fused version *) }
  simpl; reflexivity.
Defined.
\end{coqcode}
This code extracts to the optimised length function on the input list:
\begin{ocamlcode}
let rec ex2 = function | [] -> (Uint63.of_int (0))
                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
\end{ocamlcode}
